---
title: "Problem Set 9"
date: "2025-11-23"
format: html
execute: 
  eval: false
---

## Instructions

- **Limited packages**: Use only base R functions and the `dslabs` package
- **No for-loops**: Use vectorized operations and built-in functions (functions like `apply` are acceptable except where specifically noted)

---

1. Create a 100×10 matrix of randomly generated normal numbers. Store the result in `x`.

```{r}
set.seed(2025)
## your code here
```


2. Apply the three R functions that return: (a) the dimensions of `x`, (b) the number of rows of `x`, and (c) the number of columns of `x`.

```{r}
## your code here
```


3. Add row-specific scalars to matrix `x`: add 1 to row 1, add 2 to row 2, and so on.

```{r}
## your code here
```


4. Add column-specific scalars to matrix `x`: add 1 to column 1, add 2 to column 2, and so on. *Hint: Use `sweep` with `FUN = "+"`.*

```{r}
## your code here
```


5. Compute the mean of each row of `x`.

```{r}
## your code here
```


6. Compute the mean of each column of `x`.

```{r}
## your code here
```


7. Load the MNIST training data using `dslabs`. For each digit class (0-9), compute the proportion of pixels that are in the "grey area" (pixel values between 50 and 205, inclusive). Create a boxplot showing these proportions by digit class. *Hint: Use logical operators and `rowMeans`.*

```{r}
## your code here
```


8. Use the `solve` function to find the solution to this system of linear equations:

$$
\begin{align}
x + 2y - 2z &= -15\\
2x + y - 5z &= -21\\
x - 4y + z &= 18
\end{align}
$$

```{r}
## your code here
```


9. Use matrix multiplication to compute the mean of each column of `x` and store the result as a single-row matrix. *Hint: Create a 1×n matrix of weights (1/n, 1/n, ..., 1/n) where n = `nrow(x)`.*

```{r}
## your code here
```


10. Use matrix multiplication and other matrix operations to compute the standard deviation of each column of `x`. Do not use `sweep`, `apply`, or any built-in standard deviation functions. *Hint: Recall that $\text{sd} = \sqrt{\frac{1}{n-1}\sum(x_i - \bar{x})^2}$.*

```{r}
## your code here
```


11. Load the MNIST training data and create a matrix `small_x` containing only the first 100 observations. Use matrix multiplication to compute the correlation matrix between all pairs of these 100 observations. *Hint: First standardize the rows, then use `%*%` with the transpose.*

```{r}
## your code here
```


12. Using the matrix `x` from problem 1, create a new matrix where each entry is replaced by 1 if the original value is above the row mean, and 0 otherwise. Do this using only matrix operations and logical indexing (no loops or apply functions).

```{r}
## your code here
```


13. Compute the Euclidean distance between row 1 and row 50 of matrix `x` using matrix multiplication. Store the result in a variable called `distance`. *Hint: Use `crossprod()` or matrix multiplication with transpose.*

```{r}
## your code here
```


14. Create a 5×5 identity matrix using only the `matrix()` function and `diag()`. Store it in a variable called `I5`.

```{r}
## your code here
```


15. Using the MNIST data, find which pixel position (row and column in the 28×28 image grid) has the highest average intensity across all training images. Return both the row and column indices. *Hint: Use `which.max()` and `arrayInd()`.*

```{r}
## your code here
```

16. Using the MNIST training data, take the first 100 observations and reshape them into a 100×28×28 multidimensional array. Then compute the per-pixel averages across the first dimension to get a 28×28 matrix of average pixel intensities. You can visualize this average image using `image(1:28, 1:28, pixel_averages[, 28:1])` to see what the "average" digit looks like across these 100 observations.

```{r}
## your code here
```

17. Select the first 20 observations labeled as "1" and the first 20 observations labeled as "0" from the MNIST training data. Compute three correlation matrices: (a) correlations within the 1's group, (b) correlations within the 0's group, and (c) correlations between the 1's and 0's groups. Compare the average within-group correlation to the average between-group correlation.

```{r}
## your code here
```