---
title: Problem set 2 
date: 2025-09-18
---

For these exercises, do not load any packages other than **dslabs**.

Make sure to use vectorization whenever possible.

1. What is the sum of the first 100 positive integers? Use the functions `seq` and `sum` to compute the sum with R for any `n`.

```{r}
# Your code here
```


2. Load the US murders dataset from the **dslabs** package. Use the function `str` to examine the structure of the `murders` object. What are the column names used by the data frame for these five variables? Show the subset of `murders` showing states with less than 1 per 100,000 deaths. Show all variables.

```{r}
library(dslabs)
str(murders)
```

```{r}
# Your code here
```


3. Show the subset of `murders` showing states with less than 1 per 100,000 deaths and in the West of the US. Don't show the `region` variable.

```{r}
# Your code here
```


4. Show the largest state with a rate less than 1 per 100,000.

```{r}
# Your code here
```


5. Show the state with a population of more than 10 million with the lowest rate.

```{r}
# Your code here
```


6. Compute the rate for each region of the US.

```{r}
# Your code here
```


7. Create a vector of numbers that starts at 6, does not pass 55, and adds numbers in increments of 4/7: 6, 6 + 4/7, 6 + 8/7, and so on. How many numbers does the list have? Hint: use `seq` and `length`.

```{r}
# Your code here
```


8. Make this data frame:

```{r}
temp <- c(35, 88, 42, 84, 81, 30)
city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", 
          "San Juan", "Toronto")
city_temps <- data.frame(name = city, temperature = temp)
```

Convert the temperatures to Celsius.

```{r}
# Your code here
```


9. Write a function `euler` that compute the following sum for any $n$:

$$
S_n = 1+1/2^2 + 1/3^2 + \dots 1/n^2
$$

```{r}
# Your code here
```


10. Show that as $n$ gets bigger we get closer $\pi^2/6$ by plotting $S_n$ versus $n$ with a horizontal dashed line at $\pi^2/6$.

```{r}
# Your code here
```


11. Use the `%in%` operator and the predefined object `state.abb` to create a logical vector that answers the question: which of the following are actual abbreviations: MA, ME, MI, MO, MU?

```{r}
# Your code here
```


12. Extend the code you used in the previous exercise to report the one entry that is **not** an actual abbreviation. Hint: use the `!` operator, which turns `FALSE` into `TRUE` and vice versa, then `which` to obtain an index.

```{r}
# Your code here
```


13. In the `murders` dataset, use `%in%` to show all variables for New York, California, and Texas, in that order.

```{r}
# Your code here
```


14. Write a function called `vandermonde_helper` that for any $x$ and $n$, returns the vector $(1, x, x^2, x^3, \dots, x^n)$. Show the results for $x=3$ and $n=5$. 

```{r}
# Your code here
```


15. Create a vector using:

```{r}
n <- 10000
p <- 0.5
set.seed(2024-9-6)
x <- sample(c(0,1), n, prob = c(1 - p, p), replace = TRUE)
```

Compute the length of each stretch of 1s and then plot the distribution of these values. Check to see if the distribution follows a geometric distribution as the theory predicts.
Do not use a loop! 
 
```{r}
# Your code here
```

16. In the `murders` dataset, create a logical vector that indicates which states have both a murder rate higher than the national average AND a population greater than 5 million. Then use `ifelse` to create a character vector that labels these states as "High Crime, High Pop", states with murder rate higher than average but population ≤ 5 million as "High Crime, Low Pop", and all other states as "Lower Crime".

```{r}
# Your code here
```

17. Use `order`, `rank`, and `sort` functions on the murder rates to answer the following: What is the murder rate of the state that ranks 10th in terms of murder rate? Show your work by creating the murder rate vector, then using the appropriate function to find the 10th ranked value.

```{r}
# Your code here
```

18. 18. Write a function called `compute_harmonic_mean` that takes a numeric vector and returns the harmonic mean (which is $n / \sum_{i=1}^{n} 1/x_i$). The function should return `NA` if any values are zero or negative. Test your function on the vector `c(1, 2, 4, 8)` and show that it returns approximately 2.13.

```{r}
# Your code here
```

19. Create a function called `safe_divide` that takes two arguments `x` and `y` and returns their ratio `x/y`, but returns the string "Cannot divide by zero" when `y` is zero. Use vectorization so that the function works element-wise on vectors. Test it on the vectors `x <- c(10, 20, 30)` and `y <- c(2, 0, 5)`.

```{r}
# Your code here
```

20. Using the `murders` dataset, write a function called `classify_state_safety` that takes a state name as input and returns a classification based on the murder rate: "Very Safe" (rate < 1), "Safe" (rate 1-3), "Moderate" (rate 3-5), or "High Risk" (rate > 5). If the state name is not found, return "State not found". Test your function on "Vermont", "Texas", "California", and "NotAState". Then use `sapply` to classify all states and create a table showing how many states fall into each safety category using the `table` function.

```{r}
# Your code here
```