---
title: Problem set 2 (not finalized)
date: 2026-09-11
---

# Scatterplots and polynomials 

The `anscombe` dataset provided in R originates from
a [1973 paper](https://hollis.harvard.edu/permalink/01HVD_INST/726dgh/cdi_jstor_primary_10_2307_2682899).

We will come back to this dataset for various purposes.  Here
we have some inelegant plotting code.

```{r lk1}
par(mfrow=c(2,2), mar=c(4,2,1,1))
with(anscombe, plot(x1,y1,xlim=c(4,20), ylim=c(3,13)))
with(anscombe, plot(x2,y2,xlim=c(4,20), ylim=c(3,13)))
with(anscombe, plot(x3,y3,xlim=c(4,20), ylim=c(3,13)))
with(anscombe, plot(x4,y4,xlim=c(4,20), ylim=c(3,13)))
```

1. Create a markdown document "poly1.qmd".  In the yaml header, put title: poly1 and author: [your name]

2. Add the section heading `# Anscombe plots` and place the code chunk seen above in that section.

3. Add text after the plots that describes in words the purpose of the xlim and ylim settings.

4. We will now explore the role of polynomial functions in
describing patterns seen in scatterplots.

- Markdown setup

    - Add a new section heading `# A polynomial function to model the Anscombe x2:y2 data`

    - Define, in an r code chunk, the function `q1 = function(x) -7 + 2.78*x -.127*x*x`

    - Use the following commands to produce a fresh plot of the anscombe x2, y2 vectors.

```{r lk2}
par(mfrow=c(1,1))
with(anscombe, plot(x2,y2))
```

- Expand this chunk of code:

    - Produce a new vector of "x values" `newx = seq(4,14,.1)`

    - Use the lines command to add newx and q1(newx) to the x2:y2 plot

    - Define a function q2 similar to q1, for which the line traced by the values `(newx, q2(newx))` "go through" the x2:y2 points.

    - Give a mathematical explanation of the "defect" of q1 in "fitting" the x2:y2 pattern.  Describe how you "repaired" the defect.
