---
title: Problem set 2 
date: 2026-09-11
---

# Overview

This problem set addresses basic operations in R.

If you use AI to help solve any problem, please
provide information on 1) the model in use, 2) the prompt used to get help,
3) some comments on what you learned from the encounter.

Solutions to this problem set should be submitted as quarto documents.
Place the required file in your bst260 repository in folder "solutions" as "pset02.qmd"
Use the title "BST 260 Problem set 2: Basics and polynomials".  Put your
name as author.

# Review of key R concepts

## Functions

- All actions in R occur with evaluation of **functions**.

- Functions in R consist of a name, a body, and arguments.  To create a function
of one argument x that doubles its input, we can write

```{r dofunc}
doub = function(x) 2*x
doub(5)
```

Function calls can in certain cases be successfully composed.

```{r dof2}
doub(doub(5))
```

## Vectors


An ordered collection of entities of a common type (typically numeric,
character, or logical) is referred to as a "vector".
This is the simplest form of data in R. 

### Types

Vectors have "types".  The most important types are numeric, character, logical, and factor.

We will focus on numeric vectors in early problem sets.

A good overview of vectors and data types is [here](https://seandavi.github.io/RBiocBook/vectors.html);
see also [the course textbook](https://rafalab.dfci.harvard.edu/dsbook-part-1/R/R-basics.html#data-types).

### Construction

#### Functions that make vectors

- One way to create a vector is to use the `c` function, shorthand for "combine".

- Another way is with the `seq` function.  The most
commonly used arguments to `seq` are `from`, `to`, and `by`.

**Exercise 1.**  Explain the result of evaluating: `all.equal(c(1,2,3), seq(1,3,1))`

**Exercise 2.**  What is the length of the vector produced with `seq(1,100,.5)`?  Show your code.

Arithmetic operations on vectors yield new vectors.

```{r doarith}
c(1,2,3) + c(3,2,1)
```

So do logical operations:

```{r dologic}
c(1,2,3) > c(3,2,1)
```

### The recycling rule

When a binary operation (arithmetic or logical) is performed
on two vectors of unequal lengths, the shorter is "replicated"
to the length of the longer.

```{r chkrecyc}
c(1,2,3) * 10
```

We could say that this operation proceeded by copying 10 three times
and then doing element-wise multiplication.


**Exercise 3.**  Explain the result.  What is its type?
```{r lklog}
c(3,4,5,6) > 4
```


## Plotting

Base R graphics are very simple and we will start with very simple approaches,
focusing on scatterplots.  For equal-length numeric vectors x and y, plot(x,y)
produces a plot of y against x.

```{r lkpl1}
x = seq(1,10,1)
y = seq(10,1,-1)
plot(x,y)
```

We can enhance an existing plot.  For example, we can draw a line
through the points produced above.

```{r lkpl2}
plot(x,y)
lines(x,y)
```


We can control margin size or graphic layout with `par`, as seen below.


# Scatterplots and polynomials 

The `anscombe` dataset provided in R originates from
a [1973 paper](https://hollis.harvard.edu/permalink/01HVD_INST/726dgh/cdi_jstor_primary_10_2307_2682899).

We will come back to this dataset for various purposes.  Here
we have some inelegant plotting code.

```{r lk1}
par(mfrow=c(2,2), mar=c(4,2,1,1))
with(anscombe, plot(x1,y1,xlim=c(4,20), ylim=c(3,13)))
with(anscombe, plot(x2,y2,xlim=c(4,20), ylim=c(3,13)))
with(anscombe, plot(x3,y3,xlim=c(4,20), ylim=c(3,13)))
with(anscombe, plot(x4,y4,xlim=c(4,20), ylim=c(3,13)))
```

**Exercise 4.** Add a section heading `# Anscombe plots` and place the code chunk seen above in that section.

**Exercise 5.** Add text after the plots that describes in words the purpose of the xlim and ylim settings.

**Exercise 6**. We will now explore the role of polynomial functions in
describing patterns seen in scatterplots.

- Markdown setup

    - Add a new subsection heading `## A polynomial function to model the Anscombe x2:y2 data`

    - Define, in an r code chunk, the function `q1 = function(x) -7 + 2.78*x -.127*x*x`

    - Use the following commands to produce a fresh plot of the anscombe x2, y2 vectors.

```{r lk2}
par(mfrow=c(1,1))
with(anscombe, plot(x2,y2))
```

- Expand this chunk of code:

    - Produce a new vector of "x values" `newx = seq(4,14,.1)`

    - Use the `lines` command to add newx and q1(newx) to the x2:y2 plot

    - Define a function q2 similar to q1, for which the line traced by the values `(newx, q2(newx))` "go through" the x2:y2 points.

    - Give a mathematical explanation of the "defect" of q1 in "fitting" the x2:y2 pattern.  Describe how you "repaired" the defect.
